---
title: "Quarto template"
format: html
---

This is a Quarto document. It mixes together Markdown text and code and can be rendered into a variety of formats such as HTML web pages and PDFs. There are three types of content. The text at the top of the document surrounded by three dashes is the **YAML header**, which controls details for the output of the document and settings that affect the whole document. The second type of content is Markdown text; the type of text you are reading right now. Any [Markdown](https://quarto.org/docs/authoring/markdown-basics.html) styling is *open* to you to use how **you** want. The third type of content is **code chunks**, which are open and closed by three back ticks (```).

Let's make our first code chunk. It looks like the below text. You can write it out manually or use the keyboard shortcut: `Cmd + Option + I` or `Ctrl + Alt + I`.

```{r}
# Put your R code here
17 + 5
```

You can place as many or as few lines of R code in a code chunk as you want. Anything within the three back ticks acts as if it were an R script. You can run a code expression using `Ctrl + Enter` or `Cmd + Return` or run the entire chunk with the run button to the right of the chunk.

```{r}
#| label: import-data
#| message: false
library(tidyverse)
library(here)

interviews <- read_csv(here("data-raw", "SAFI_clean.csv"), na = "NULL")
```

You will notice something a little different about the comments in the above code chunk. Instead of a comment, `#|` sets code chunk options in the form of `option: value` just like in the YAML at the top of the document. The above chunk includes a label and tells Quarto not to output messages, such as the startup message when you load the tidyverse package. See 

```{r}
#| echo: false
#| fig-width: 6
#| fig-asp: 0.618
#| fig-align: center

ggplot(data = interviews) + 
  geom_bar(aes(x = respondent_wall_type, fill = village), position = "dodge") + 
  scale_fill_manual(values = c("darkorange", "purple", "cyan4")) + 
  labs(title = "Studying African Farmer-Led Irrigation",
       x = "Household wall type",
       y = NULL,
       fill = "Village") + 
  theme_minimal()
```

You can make data visualizations and have their output included right in the document. With the option `#| echo: false` only the plot will be output, not the code that created it. Click the **Render** button at the top of script pane in RStudio or use or press `Cmd/Ctrl + Shift + K` to see what the output looks like.